Global code motion of congruent instructions

### 2017-04-18 paper.tex
ispell-buffer

We have implemented the pass in \LLVM{}. It is based on Global Value Numbering
infrastructure available in \LLVM{}. The experimental results show an average
saving of 1\% on overall compile time on \SPEC{}. \GCM{} enables more inlining
and exposes more loop invariant code motion opportunities in majority of the
benchmarks. We have also seen execution time improvements in a few of \SPEC{}
benchmarks viz. mcf and sjeng, moreover, register spills reduced by 2\% on the
\SPEC{} bencmarks suite when compiled for \xlinux{}. This is an optimistic
algorithm in the sense that it considers all congruent instructions in a
function as potential candidates. We make an extra effort to hoist candidates by
partitioning the potential candidates in a way to enable partial hoisting in
case common hoisting points for all the candidates cannot be found. We also
formalize how register pressure will reduce as a result of \gcm{}, and how
\gcm{} increases instruction level parallelism thereby enabling more \ooo{}
execution on modern architectures.
\end{abstract}

@@

%% Keywords
%% comma separated list
\keywords{Optimizng Compilers, \GCC{}, \LLVM{}, Code Generation, Global Scheduling}  %% \keywords is optional

@@

In this code the division operation can be executed in parallel with the two
subtractions because there are no dependencies among them. So the total number
of cycles will be $max(C_{div}, C_{sub}) + C_{mul} = C_{div} + C_{mul}$; since
$C_{div}$ is usually much greater than $C_{sub}$ \cite{x86,aarch64}. We have
seen the performance of an \ooo{} processor going up by $15\%$ on a benchmark
containing the above motivating example running in a tight loop.  In fact, there
are several advantages of \GCM{}:
\begin{itemize}
\item it helps reduce the code size of the program by replacing multiple
  instructions with one although the final code size may be higher because of
  improved inlining heuristics; functions become cheaper to inline when their
  code size decreases because the inliner heuristics depend on instruction
  count;
\item it exposes more instruction level parallelism to the later compiler
  passes. By hoisting identical computations to be executed earlier, instruction
  schedulers can move heavy computations earlier in order to avoid pipeline
  bubbles;
\item it reduces branch misprediction penalty on \ooo{} processors with
  speculative execution of branches: by hoisting or sinking expressions out of
  branches, it can effectively reduce the amount of code to be speculatively
  executed and hence reduce the critical path;
\item it reduces interference or register pressure with an appropriate
  cost-model: we have used a cost model in our implementation which only hoists
  or sinks when the register pressure is reduced;
\item it reduces the total number of instructions to be executed for SIMD
  architectures which execute all code in branches based on masking or
  predication;
\item it may improve loop vectorization by reducing a loop with control flow to
  a loop with a single basic-block, should all the instructions in a conditional get
  hoisted or sinked;
\item it enables more loop invariant code motion (LICM): as LICM passes, in
  general, cannot effectively reason about instructions within conditional
  branches in the context of loops, code-hoisting is needed to move instructions
  out of conditional expressions and expose them to LICM.
\end{itemize}

@@

There are a lot of bug reports in \GCC{} and \LLVM{} bugzillas
\cite{GCCCodeHoistingBugs, LLVMCodeHoistingBugs}, showing the interest in having
a more powerful code hoist transform.  The current \LLVM{} implementation of
code hoisting in SimplifyCFG.cpp is very limited to hoisting from identical
basic-blocks: the instructions of two sibling basic-blocks are read at the same
time, and all the instructions of the blocks are hoisted to the common parent
block as long as the compiler is able to prove that the instructions are
equivalent.  This implementation does not allow for an easy extension: first in
terms of compilation time overhead the implementation is quadratic in number of
instructions to bisimulate and second, the equivalence of instructions is
computed by comparing the operands which is neither general nor scalabale.

@@

\citet{shobaki2013} also recently presented a combinatorial global
scheduler with reasonable performance improvements. It is possible that both
Barany and Shobaki's implmentation will have similar results when compiled for
same target architecture. Also, both suffer from the same problem, although
Shobaki not so much, of large compile times which would not suit in industrial
compilers like \GCC{} and \LLVM{}. With the algorithm described in the next
section, we got a reduction in register spills, improved inliner heuristics,
improved compile time, and show performance improvements on \SPEC{} benchmarks.

@@

We compute merge sets, of the control flow graph, which remain same throughout
the \gcm{} transformation. For \gcm{} we only want to know if a use
operand is a kill (to compute changes in register pressure.) For that we only
need to know whether the use is also required later in the execution path. A
variable is live out of a basic-block B if it is used in the merge set of B. We
compute the live-out relation on-demand when profitability of hoistable/sinkable
candidates is to be evaluated. A simplified version of
\emph{isLiveOutUsingMergeSet} is presented here:

@@

\subsubsection{Legality of hoisting loads}
The availability of operand to the load (an address) is checked at the hoisting
point. If that is not available we try to rematerialize the addresss if
possible.  Along the path, from current position of the load instruction
backwards on the control flow to the hoisting point, we check whether there are
writes to memory that may alias with the load, in which case the candidate is
discarded. To iterate on the use-def chains for memory references the MemorySSA
infrastructure of \LLVM{} is used.

@@

In this example (c1, c2) or (d1, d2) are potential sinkable candidates. Since
(c1, c2) depend on i1 and i2 respectively which are also in their original basic
blocks, c1 and c2 are not anticipable in B3. So without knowing the sinkability
of `i1' and `i2' it would be illegal to sink (c1, c2) to B3. On the other hand
(d1, d2) can safely be sunk because their operands are readily available at the
sink point, i.e., B3. It should also be noted that, just because the expressions
are identical and operands are available, it still requires a unique
post-dominating PHI to use the exact same values to be legally sinkable.

A general global scheduling algorithm also requires checks for undefinedness
when introducing a new computation along a path. Since only very busy
expressions are moved in the current implementation, there is no need to check
for undefinedness resulting due to movement of instructions. This simplifies the
implementation.

@@

Ideally, it should be okay to hoist all the instructions and a later
live-range-splitting \cite{cooper1998live} pass should make the right decision
of rematerializing the instruction should it be beneficial to do so. But the
current live-range splitting pass of \LLVM{} is not making the optimial decision
and we have found performance regressions while hoisting aggressively.

@@

\subsubsection{Partitioning the list of hoistable candidates to maximize hoisting}
\label{subsec:partition}
%The partition algorithm is described as follows:
In the approach, described in Section~\ref{subsec:optimistic}, it is possible
that a common hoisting point for all the instructions does not satisfy legality
or profitability checks. In these cases, it is still possible to `partially'
hoist a subset of instructions by splitting the set of candidates and finding a
closer hoisting point for each subset.

In order to hoist a subset of identical instructions, we partition the list of
all candidates in a way to maximize the total number of hoistings.  By sorting
the list of all the candidates in the increasing order of their depth first
search discovery time stamp \cite{clrs} (DFSIn numbers), we make sure that
candidates closer in the list have their common dominator nearby. In
Figure-\ref{fig:dfsin}, if B3, B4, B5, and B6 have identical computations and if
for some reason, they cannot be hoisted at B0, then we partition the set of
hoistable candidates in their DFSIn order. In this case the DFSIn ordering would
be B3, B4, B5, B6 which will allow the instructions in (B3, B4) to be hoisted at
B1 and those in (B5, B6) to be hoisted at B2. Even if instructions in (B3, B4)
could not be hoisted, for some reason, to B1, the instructions in (B5, B6) can
still be hoisted -- if legal and profitable -- to B2.

@@

After one iteration of algorithm runs through the entire function, it creates
more oppportunities for \emph{higher ranked} computations
\cite{rosen1988global}. Currently, this is a limitation of the \GVN{} analysis
pass, and so we rerun the code-hoisting algorithm until there are no more
instructions left to be hoisted.  Obviously, this is not the most optimal
approach and can be improved by ranking the computations \cite{rosen1988global},
or by improving the \GVN{} analysis to correctly populate congruence classes as
the program is modified by the code generation.

@@

We collect the \GVN{} of all the instructions in the function and iterate on the
list of instructions having identical \GVN{}s. The algorithm prefers hoisting to
sinking. So first we find the common dominator dominating all such identical
computations and perform legality checks, as described in
Section~\ref{subsec:legality}. Often times it is not possible to hoist all the
instructions to one common dominator, due to legality constraints, e.g.,
intersecting side-effects, or profitability constraints, e.g., increasing
register pressure; in those cases, this algorithmn would partition
(Section~\ref{subsec:partition}) the list of identical instructions into subsets
which can be partially hoisted to their respective common dominators. However,
it should be noted that we only hoist very busy expressions to avoid checking
for undefined behaviors resulting because of introducing extra computation in an
execution path.

@@

\newpage
After no more candidates are hoistable, sinking is performed. Sinking is only
performed once becase currently there are very few sinkable candidates
(Table~\ref{tab:code-motion-metric}) as SimplifyCFG of LLVM (which runs before
\GCM{}) has a code-sinking which already sinks several instructions.

@@

Code hoisting opens new opportunities for other hoistable candidates which were
of higher rank (depended on candidates which got hoisted). Ideally we could
iterate on lower ranking expressions first and then proceed to higher ranking
expressions in the same iteration but \LLVM{'}s \GVN{} infrastucture does not
compute equivalence classes in an effective way. We found it simpler to just
recompute the value numbers and start finding hoistable candidates again.

\subsection{Time complexity of algorithm}
The complexity of code hoisting is linear in number of instructions that are
candidates for \gcm{}, matching the complexity of \PRE{} on \SSA{} form.
The analyses computed for this pass are Global Value Numbering, Computation of
DJ-graph and merge-sets, Marking Barriers, all are linear in number of
instructions in a function. Liveness analysis is expensive and only performed
on-demand for hoistable candidates so it does not affect the overall compilation
time by much. Other analyses like Alias Analysis, \MemorySSA{} and Dominator Tree
are already available in the \LLVM{} pass pipeline. Although we recompute \GVN{}, and
Barriers for each iteration of the \gcm{}, we have not seen signinficant
increase in the compilation times (see Section~\ref{tab:compile-time}). We have also
provided appropriate compiler
flags to expedite \gcm{} by bailing out with fewer iterations, or skip the
liveness based profitability analysis to aggressively move the code as long as
they are legal.

@@

\begin{table}[h!]
  \begin{center}
    \begin{tabular}{|l|c|c|c|}
      \hline
% TODO: Check if putting code-size or spills number is okay or only %age is required.
% llvm trunk vs. llvm trunk with code-motion
                &\% performance uplift & \% codesize decrease & \% codesize decrease\\
Benchmarks      & at -Ofast (high better)& at -Ofast (low better) & at -Os (low better)\\\hline
400.perlbench	& 1.00  & 1.00 & 1.00\\\hline
401.bzip2	& 1.00  & 1.00 & 1.00\\\hline
403.gcc	        & 1.00  & 1.01 & 1.00\\\hline
429.mcf	        & 1.03  & 1.00 & 1.00\\\hline
433.milc	& 1.00  & 1.00 & 1.00\\\hline
444.namd	& 1.00  & 1.01 & 1.00\\\hline
445.gobmk	& 0.99  & 1.01 & 1.00\\\hline
447.dealII	& 0.95  & 1.01 & 0.98\\\hline
450.soplex	& 0.99  & 0.99 & 1.00\\\hline
453.povray	& 0.97  & 1.00 & 1.00\\\hline
456.hmmer	& 0.99  & 1.00 & 1.00\\\hline
458.sjeng	& 1.02  & 1.00 & 1.00\\\hline
462.libquantum	& 1.00  & 1.00 & 1.00\\\hline
464.h264ref	& 1.00  & 1.00 & 1.00\\\hline
470.lbm	        & 1.00  & 1.00 & 1.00\\\hline
471.omnetpp	& 1.00  & 0.99 & 1.00\\\hline
473.astar	& 1.00  & 1.02 & 1.00\\\hline
482.sphinx3	& 0.99  & 1.00 & 1.00\\\hline
483.xalancbmk	& 1.01  & 0.99 & 1.00\\\hline
Geomean         & 0.997 & 1.00 & 1.00\\\hline
    \end{tabular}
  \end{center}
  \caption{Execution time and code size with and without \gcm{}
    (\GCM{}) on \SPEC{}: performance uplift at -Ofast in mcf and sjeng; code size
    improvement in dealII at -Os.}
  \label{tab:code-size}
\end{table}

Overall there was only a minor change in the performance numbers with a few improvements 
(429.mcf, 458.sjeng etc.) and a few regresssions (447.dealII, 453.povray
etc.). The \gcm{} pass was run twice in the pass pipeline (at -Ofast),
first after EarlyCSE pass, and then after the inliner. Because EarlyCSE removes
local redundancies, \GCM{} would have to analyze less redundant instructions, also
as inliner creates more redundancies, \GCM{} would have more opportunities to move
congruent instructions out of branches.

Code size also remained almost the same as a result of \gcm{} as shown in
Table~\ref{tab:code-size}.  The code size listed here is the text size (from
linux size command) of the final executables for each benchmark.  Some bencmarks
like 473.astar increased in code size by almost 2\%. This is because once the
code-size of a function decreases (due to \GCM{}), it becomes cheaper to
inline. As we can see in Table-\ref{tab:stats}, four more functions got inlined
in 473.astar. Except for 447.dealII, 450.soplex, 482.sphinx3 and 483.xalancbmk
all other benchmarks got more functions inlined. Also, since the pass runs
early, it affects many optimizations which rely on the number of instructions,
length of the use-def chain, and other metrics.

@@

There are a few improvements that we would like to make. First would be to
integrate this with \LLVM{'}s new GVN-PRE implementation such that more
congruent instructions can be scheduled. This would also expose more
redundancies to the GVN-PRE.  In general it is a good idea to start with lower
ranked expressions first such that maximum hoisting can happen in one iteration,
however, current implementation does not rank the expressions and iteratively
finds a fixed point when no more candidates are available. Even this
implementation converges quickly and no compile time regression have been
observed. This is not the most optimal approach in terms of compile time complexity
and results in multiple data structures to be recomputed. This can be improved
by ranking the computations \cite{rosen1988global}. Also, since \GCM{} runs very
early in the pass pipeline, it will be good to evaluate the codesize/performance
impact when it is run in sync with \GVN{}-\PRE{} just like \GCC{} does.

>>>>

We have implemented the pass in \LLVM{}. It is based on Global Value Numbering
infrastructure available in \LLVM{}. The experimental results show an average
saving of 1\% on overall compile time on \SPEC{}. \GCM{} enables more inlining
and exposes more loop invariant code motion opportunities in majority of the
benchmarks. We have also seen execution time improvements in a few of \SPEC{}
benchmarks viz. mcf and sjeng, moreover, register spills reduced by 2\% on the
\SPEC{} benchmarks suite when compiled for \xlinux{}. This is an optimistic
algorithm in the sense that it considers all congruent instructions in a
function as potential candidates. We make an extra effort to hoist candidates by
partitioning the potential candidates in a way to enable partial hoisting in
case common hoisting points for all the candidates cannot be found. We also
formalize how register pressure will reduce as a result of \gcm{}, and how
\gcm{} increases instruction level parallelism thereby enabling more \ooo{}
execution on modern architectures.
\end{abstract}

@@

%% Keywords
%% comma separated list
\keywords{Optimizing Compilers, \GCC{}, \LLVM{}, Code Generation, Global Scheduling}  %% \keywords is optional

@@

In this code the division operation can be executed in parallel with the two
subtractions because there are no dependencies among them. So the total number
of cycles will be $max(C_{div}, C_{sub}) + C_{mul} = C_{div} + C_{mul}$; since
$C_{div}$ is usually much greater than $C_{sub}$ \cite{x86,aarch64}. We have
seen the performance of an \ooo{} processor going up by $15\%$ on a benchmark
containing the above motivating example running in a tight loop.  In fact, there
are several advantages of \GCM{}:
\begin{itemize}
\item it helps reduce the code size of the program by replacing multiple
  instructions with one although the final code size may be higher because of
  improved inlining heuristics; functions become cheaper to inline when their
  code size decreases because the inliner heuristics depend on instruction
  count;
\item it exposes more instruction level parallelism to the later compiler
  passes. By hoisting identical computations to be executed earlier, instruction
  schedulers can move heavy computations earlier in order to avoid pipeline
  bubbles;
\item it reduces branch misprediction penalty on \ooo{} processors with
  speculative execution of branches: by hoisting or sinking expressions out of
  branches, it can effectively reduce the amount of code to be speculatively
  executed and hence reduce the critical path;
\item it reduces interference or register pressure with an appropriate
  cost-model: we have used a cost model in our implementation which only hoists
  or sinks when the register pressure is reduced;
\item it reduces the total number of instructions to be executed for SIMD
  architectures which execute all code in branches based on masking or
  predication;
\item it may improve loop vectorization by reducing a loop with control flow to
  a loop with a single basic-block, should all the instructions in a conditional get
  hoisted or sunk;
\item it enables more loop invariant code motion (LICM): as LICM passes, in
  general, cannot effectively reason about instructions within conditional
  branches in the context of loops, code-hoisting is needed to move instructions
  out of conditional expressions and expose them to LICM.
\end{itemize}

@@

There are a lot of bug reports in \GCC{} and \LLVM{} bugzillas
\cite{GCCCodeHoistingBugs, LLVMCodeHoistingBugs}, showing the interest in having
a more powerful code hoist transform.  The current \LLVM{} implementation of
code hoisting in SimplifyCFG.cpp is very limited to hoisting from identical
basic-blocks: the instructions of two sibling basic-blocks are read at the same
time, and all the instructions of the blocks are hoisted to the common parent
block as long as the compiler is able to prove that the instructions are
equivalent.  This implementation does not allow for an easy extension: first in
terms of compilation time overhead the implementation is quadratic in number of
instructions to bisimulate and second, the equivalence of instructions is
computed by comparing the operands which is neither general nor scalable.

@@

\citet{shobaki2013} also recently presented a combinatorial global
scheduler with reasonable performance improvements. It is possible that both
Barany and Shobaki's implementation will have similar results when compiled for
same target architecture. Also, both suffer from the same problem, although
Shobaki not so much, of large compile times which would not suit in industrial
compilers like \GCC{} and \LLVM{}. With the algorithm described in the next
section, we got a reduction in register spills, improved inliner heuristics,
improved compile time, and show performance improvements on \SPEC{} benchmarks.

@@

We compute merge sets, of the control flow graph, which remain same throughout
the \gcm{} transformation. For \gcm{} we only want to know if a use
operand is a kill (to compute changes in register pressure.) For that we only
need to know whether the use is also required later in the execution path. A
variable is live out of a basic-block B if it is used in the merge set of B. We
compute the live-out relation on-demand when profitability of hoist/sink
candidates is to be evaluated. A simplified version of
\emph{isLiveOutUsingMergeSet} is presented here:

@@

\subsubsection{Legality of hoisting loads}
The availability of operand to the load (an address) is checked at the hoisting
point. If that is not available we try to rematerialize the address if
possible.  Along the path, from current position of the load instruction
backwards on the control flow to the hoisting point, we check whether there are
writes to memory that may alias with the load, in which case the candidate is
discarded. To iterate on the use-def chains for memory references the MemorySSA
infrastructure of \LLVM{} is used.

@@

In this example (c1, c2) or (d1, d2) are potential sinkable candidates. Since
(c1, c2) depend on i1 and i2 respectively which are also in their original basic
blocks, c1 and c2 are not anticipable in B3. So without knowing the ability to sink
of `i1' and `i2' it would be illegal to sink (c1, c2) to B3. On the other hand
(d1, d2) can safely be sunk because their operands are readily available at the
sink point, i.e., B3. It should also be noted that, just because the expressions
are identical and operands are available, it still requires a unique
post-dominating PHI to use the exact same values to be legally sinkable.

A general global scheduling algorithm also requires checks for undefined values
when introducing a new computation along a path. Since only very busy
expressions are moved in the current implementation, there is no need to check
for undefined resulting due to movement of instructions. This simplifies the
implementation.

@@

Ideally, it should be okay to hoist all the instructions and a later
live-range-splitting \cite{cooper1998live} pass should make the right decision
of rematerializing the instruction should it be beneficial to do so. But the
current live-range splitting pass of \LLVM{} is not making the optimal decision
and we have found performance regressions while hoisting aggressively.

@@

\subsubsection{Partitioning the list of hoist candidates to maximize hoisting}
\label{subsec:partition}
%The partition algorithm is described as follows:
In the approach, described in Section~\ref{subsec:optimistic}, it is possible
that a common hoisting point for all the instructions does not satisfy legality
or profitability checks. In these cases, it is still possible to `partially'
hoist a subset of instructions by splitting the set of candidates and finding a
closer hoisting point for each subset.

In order to hoist a subset of identical instructions, we partition the list of
all candidates in a way to maximize the total number of hoist instructions.  By sorting
the list of all the candidates in the increasing order of their depth first
search discovery time stamp \cite{clrs} (DFSIn numbers), we make sure that
candidates closer in the list have their common dominator nearby. In
Figure-\ref{fig:dfsin}, if B3, B4, B5, and B6 have identical computations and if
for some reason, they cannot be hoisted at B0, then we partition the set of
hoistable candidates in their DFSIn order. In this case the DFSIn ordering would
be B3, B4, B5, B6 which will allow the instructions in (B3, B4) to be hoisted at
B1 and those in (B5, B6) to be hoisted at B2. Even if instructions in (B3, B4)
could not be hoisted, for some reason, to B1, the instructions in (B5, B6) can
still be hoisted -- if legal and profitable -- to B2.

@@

After one iteration of algorithm runs through the entire function, it creates
more opportunities for \emph{higher ranked} computations
\cite{rosen1988global}. Currently, this is a limitation of the \GVN{} analysis
pass, and so we rerun the code-hoisting algorithm until there are no more
instructions left to be hoisted.  Obviously, this is not the most optimal
approach and can be improved by ranking the computations \cite{rosen1988global},
or by improving the \GVN{} analysis to correctly populate congruence classes as
the program is modified by the code generation.

@@

We collect the \GVN{} of all the instructions in the function and iterate on the
list of instructions having identical \GVN{}s. The algorithm prefers hoisting to
sinking. So first we find the common dominator dominating all such identical
computations and perform legality checks, as described in
Section~\ref{subsec:legality}. Often times it is not possible to hoist all the
instructions to one common dominator, due to legality constraints, e.g.,
intersecting side-effects, or profitability constraints, e.g., increasing
register pressure; in those cases, this algorithm would partition
(Section~\ref{subsec:partition}) the list of identical instructions into subsets
which can be partially hoisted to their respective common dominators. However,
it should be noted that we only hoist very busy expressions to avoid checking
for undefined behaviors resulting because of introducing extra computation in an
execution path.

@@

\newpage
After no more candidates are hoistable, sinking is performed. Sinking is only
performed once because currently there are very few sinkable candidates
(Table~\ref{tab:code-motion-metric}) as SimplifyCFG of LLVM (which runs before
\GCM{}) has a code-sinking which already sinks several instructions.

@@

Code hoisting opens new opportunities for other hoistable candidates which were
of higher rank (depended on candidates which got hoisted). Ideally we could
iterate on lower ranking expressions first and then proceed to higher ranking
expressions in the same iteration but \LLVM{'}s \GVN{} infrastructure does not
compute equivalence classes in an effective way. We found it simpler to just
recompute the value numbers and start finding hoistable candidates again.

\subsection{Time complexity of algorithm}
The complexity of code hoisting is linear in number of instructions that are
candidates for \gcm{}, matching the complexity of \PRE{} on \SSA{} form.
The analyses computed for this pass are Global Value Numbering, Computation of
DJ-graph and merge-sets, Marking Barriers, all are linear in number of
instructions in a function. Liveness analysis is expensive and only performed
on-demand for hoistable candidates so it does not affect the overall compilation
time by much. Other analyses like Alias Analysis, \MemorySSA{} and Dominator Tree
are already available in the \LLVM{} pass pipeline. Although we recompute \GVN{}, and
Barriers for each iteration of the \gcm{}, we have not seen significant
increase in the compilation times (see Section~\ref{tab:compile-time}). We have also
provided appropriate compiler
flags to expedite \gcm{} by bailing out with fewer iterations, or skip the
liveness based profitability analysis to aggressively move the code as long as
they are legal.

@@

\begin{table}[h!]
  \begin{center}
    \begin{tabular}{|l|c|c|c|}
      \hline
% TODO: Check if putting code-size or spills number is okay or only %age is required.
% llvm trunk vs. llvm trunk with code-motion
                &\% performance uplift & \% code size decrease & \% code size decrease\\
Benchmarks      & at -Ofast (high better)& at -Ofast (low better) & at -Os (low better)\\\hline
400.perlbench	& 1.00  & 1.00 & 1.00\\\hline
401.bzip2	& 1.00  & 1.00 & 1.00\\\hline
403.gcc	        & 1.00  & 1.01 & 1.00\\\hline
429.mcf	        & 1.03  & 1.00 & 1.00\\\hline
433.milc	& 1.00  & 1.00 & 1.00\\\hline
444.namd	& 1.00  & 1.01 & 1.00\\\hline
445.gobmk	& 0.99  & 1.01 & 1.00\\\hline
447.dealII	& 0.95  & 1.01 & 0.98\\\hline
450.soplex	& 0.99  & 0.99 & 1.00\\\hline
453.povray	& 0.97  & 1.00 & 1.00\\\hline
456.hmmer	& 0.99  & 1.00 & 1.00\\\hline
458.sjeng	& 1.02  & 1.00 & 1.00\\\hline
462.libquantum	& 1.00  & 1.00 & 1.00\\\hline
464.h264ref	& 1.00  & 1.00 & 1.00\\\hline
470.lbm	        & 1.00  & 1.00 & 1.00\\\hline
471.omnetpp	& 1.00  & 0.99 & 1.00\\\hline
473.astar	& 1.00  & 1.02 & 1.00\\\hline
482.sphinx3	& 0.99  & 1.00 & 1.00\\\hline
483.xalancbmk	& 1.01  & 0.99 & 1.00\\\hline
Geomean         & 0.997 & 1.00 & 1.00\\\hline
    \end{tabular}
  \end{center}
  \caption{Execution time and code size with and without \gcm{}
    (\GCM{}) on \SPEC{}: performance uplift at -Ofast in mcf and sjeng; code size
    improvement in dealII at -Os.}
  \label{tab:code-size}
\end{table}

Overall there was only a minor change in the performance numbers with a few improvements 
(429.mcf, 458.sjeng etc.) and a few regressions (447.dealII, 453.povray
etc.). The \gcm{} pass was run twice in the pass pipeline (at -Ofast),
first after EarlyCSE pass, and then after the inliner. Because EarlyCSE removes
local redundancies, \GCM{} would have to analyze less redundant instructions, also
as inliner creates more redundancies, \GCM{} would have more opportunities to move
congruent instructions out of branches.

Code size also remained almost the same as a result of \gcm{} as shown in
Table~\ref{tab:code-size}.  The code size listed here is the text size (from
linux size command) of the final executable for each benchmark.  Some benchmarks
like 473.astar increased in code size by almost 2\%. This is because once the
code-size of a function decreases (due to \GCM{}), it becomes cheaper to
inline. As we can see in Table-\ref{tab:stats}, four more functions got inlined
in 473.astar. Except for 447.dealII, 450.soplex, 482.sphinx3 and 483.xalancbmk
all other benchmarks got more functions inlined. Also, since the pass runs
early, it affects many optimizations which rely on the number of instructions,
length of the use-def chain, and other metrics.

@@

There are a few improvements that we would like to make. First would be to
integrate this with \LLVM{'}s new GVN-PRE implementation such that more
congruent instructions can be scheduled. This would also expose more
redundancies to the GVN-PRE.  In general it is a good idea to start with lower
ranked expressions first such that maximum hoisting can happen in one iteration,
however, current implementation does not rank the expressions and iteratively
finds a fixed point when no more candidates are available. Even this
implementation converges quickly and no compile time regression have been
observed. This is not the most optimal approach in terms of compile time complexity
and results in multiple data structures to be recomputed. This can be improved
by ranking the computations \cite{rosen1988global}. Also, since \GCM{} runs very
early in the pass pipeline, it will be good to evaluate the code size/performance
impact when it is run in sync with \GVN{}-\PRE{} just like \GCC{} does.
